---
title: "Pips performance report"
author: "David Jenkins"
date: today
format: 
  html:
    theme: default
    css: nyt-games.css
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Overview

Compare David's and Ryan's solve times, consistency, and recorded wins for Easy, Medium, and Hard puzzles. The charts describe recorded results; they don't explain why performance changed.

```{r setup}
if (file.exists(".Renviron")) {
  readRenviron(".Renviron")
}

# Load required libraries
library(tidyverse)
library(googlesheets4)
library(lubridate)
library(DT)
library(zoo)
library(RColorBrewer)

# Authenticate and load data. Credentials are stored in the project-local
# gargle cache configured by .Renviron; run auth_google.R once to create it.
sheet_url <- "https://docs.google.com/spreadsheets/d/1DqFP8_LM0bnuNjXKArOCDJyFjFYqv4zrOqKz3RG2sMU/edit?gid=0#gid=0"
auth_cache <- Sys.getenv("GARGLE_OAUTH_CACHE", unset = ".secrets/gargle")
auth_email <- Sys.getenv("PIP_ANALYTICS_GOOGLE_EMAIL", unset = "")

if (!dir.exists(auth_cache)) {
  dir.create(auth_cache, recursive = TRUE, showWarnings = FALSE)
}

gs4_auth(
  email = if (nzchar(auth_email)) auth_email else TRUE,
  scopes = "https://www.googleapis.com/auth/spreadsheets.readonly",
  cache = auth_cache
)

raw_grid <- read_sheet(sheet_url, sheet = "Times", col_names = FALSE, range = "A:M")

cell_value <- function(column, row) {
  value <- raw_grid[[column]][[row]]

  if (length(value) == 0) {
    return(NA)
  }

  value[[1]]
}

date_value <- function(row) {
  value <- cell_value(1, row)

  if (inherits(value, c("Date", "POSIXct", "POSIXt"))) {
    return(as_date(value))
  }

  as_date(mdy(value))
}

time_seconds <- function(column, row) {
  value <- cell_value(column, row)

  if (length(value) == 0 || is.na(value)) {
    return(NA_real_)
  }

  if (inherits(value, c("POSIXct", "POSIXt"))) {
    return(hour(value) * 3600 + minute(value) * 60 + second(value))
  }

  if (is.numeric(value)) {
    return(value * 24 * 60 * 60)
  }

  hms_value <- hms(value)
  hour(hms_value) * 3600 + minute(hms_value) * 60 + second(hms_value)
}

winner_value <- function(column, row) {
  value <- cell_value(column, row)

  if (length(value) == 0 || is.na(value)) {
    return(NA_character_)
  }

  as.character(value)
}

data_rows <- seq(3, nrow(raw_grid))
time_columns <- tribble(
  ~Player, ~Difficulty, ~Column,
  "Ryan", "Easy", 2,
  "Ryan", "Medium", 3,
  "Ryan", "Hard", 4,
  "David", "Easy", 5,
  "David", "Medium", 6,
  "David", "Hard", 7
)

winner_columns <- tribble(
  ~Difficulty, ~Column,
  "Easy", 8,
  "Medium", 9,
  "Hard", 10,
  "Cookies", 11,
  "Time", 12,
  "Wins", 13
)

winner_data <- winner_columns %>%
  pmap_dfr(function(Difficulty, Column) {
    tibble(
      Date = map(data_rows, date_value) %>% list_c(),
      Difficulty = Difficulty,
      Winner = map_chr(data_rows, ~ winner_value(Column, .x))
    )
  }) %>%
  filter(!is.na(Date), !is.na(Winner))

overall_winners <- winner_data %>%
  filter(Difficulty == "Wins") %>%
  select(Date, Winner)

difficulty_levels <- c("Easy", "Medium", "Hard")
hard_time_cap_minutes <- 20
set1_colors <- brewer.pal(9, "Set1")
difficulty_colors <- setNames(set1_colors[seq_along(difficulty_levels)], difficulty_levels)
minute_breaks <- c(0.1, 0.25, 0.5, 1, 2, 5, 10, 20, 60)
nyt_black <- "#121212"
nyt_gray <- "#5a5a5a"
nyt_grid <- "#dfdfdf"
pips_purple <- "#9251ca"
pips_soft_pink <- "#d3aace"
mobile_title_width <- 25
mobile_subtitle_width <- 42

chart_title <- function(title, title_width = Inf) {
  if (is.finite(title_width)) {
    return(str_wrap(title, width = title_width))
  }

  title
}

theme_nyt_games <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "Arial") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = nyt_grid, linewidth = 0.35),
      panel.grid.minor = element_line(color = "#ebebeb", linewidth = 0.25),
      plot.title = element_text(face = "bold", size = 18, color = nyt_black, lineheight = 0.95),
      plot.subtitle = element_text(size = 11, color = nyt_gray, margin = margin(b = 12)),
      axis.title = element_text(face = "bold", color = nyt_black),
      axis.text = element_text(color = nyt_gray),
      strip.text = element_text(face = "bold", color = nyt_black, size = 12),
      legend.title = element_text(face = "bold", color = nyt_black),
      legend.text = element_text(color = nyt_black),
      plot.margin = margin(12, 12, 12, 12)
    )
}

# Data Cleaning
pips_data <- time_columns %>%
  pmap_dfr(function(Player, Difficulty, Column) {
    tibble(
      Date = map(data_rows, date_value) %>% list_c(),
      Player = Player,
      Difficulty = Difficulty,
      Solve_Time_Seconds = map_dbl(data_rows, ~ time_seconds(Column, .x))
    )
  }) %>%
  left_join(
    winner_data %>% filter(Difficulty %in% c("Easy", "Medium", "Hard")),
    by = c("Date", "Difficulty")
  ) %>%
  mutate(
    Difficulty = factor(Difficulty, levels = difficulty_levels),
    Recorded_Solve_Time_Seconds = as.numeric(Solve_Time_Seconds),
    Solve_Time_Seconds = if_else(
      Difficulty == "Hard",
      pmin(Recorded_Solve_Time_Seconds, hard_time_cap_minutes * 60),
      Recorded_Solve_Time_Seconds
    ),
    Solve_Time_Minutes = Solve_Time_Seconds / 60,
    Month = floor_date(Date, "month")
  ) %>%
  # Filter out games that don't have a recorded time
  filter(!is.na(Solve_Time_Seconds))

scale_month_axis <- function() {
  scale_x_date(
    breaks = seq.Date(min(pips_data$Month), max(pips_data$Month), by = "month"),
    date_labels = "%b\n%y"
  )
}

theme_month_axis <- function(is_mobile = FALSE) {
  theme(axis.text.x = element_text(size = if (is_mobile) 9 else 10, lineheight = 0.9))
}

monthly_trend_data <- pips_data %>%
  group_by(Player, Difficulty, Month) %>%
  summarise(
    Median_Minutes = median(Solve_Time_Minutes, na.rm = TRUE),
    Solves = n(),
    Last_Date = max(Date),
    .groups = "drop"
  ) %>%
  arrange(Player, Difficulty, Month) %>%
  group_by(Player, Difficulty) %>%
  mutate(
    Previous_Month = lag(Month),
    Previous_Median = if_else(
      Previous_Month == Month %m-% months(1),
      lag(Median_Minutes),
      NA_real_
    ),
    Change_Minutes = Median_Minutes - Previous_Median,
    Change_Percent = if_else(
      Previous_Median > 0,
      Change_Minutes / Previous_Median,
      NA_real_
    ),
    Is_Partial = Month >= floor_date(today(tzone = "America/New_York"), "month")
  ) %>%
  ungroup()

report_month <- floor_date(today(tzone = "America/New_York"), "month")

format_minutes <- function(value, accuracy = 0.01) {
  ifelse(is.na(value), "Not available", paste(scales::number(value, accuracy = accuracy), "min"))
}

pace_change <- function(change) {
  case_when(
    is.na(change) ~ "No comparison available",
    near(change, 0) ~ "Unchanged",
    change < 0 ~ paste(format_minutes(abs(change), 0.001), "faster"),
    TRUE ~ paste(format_minutes(abs(change), 0.001), "slower")
  )
}

summary_cell <- function(headline, ...) {
  htmltools::tagList(
    htmltools::tags$p(htmltools::tags$strong(headline)),
    lapply(list(...), htmltools::tags$p)
  )
}

summary_table <- function(headers, rows, caption, statistical = FALSE) {
  htmltools::tags$div(
    class = "report-summary",
    htmltools::tags$table(
      class = if (statistical) "summary-table statistical-summary" else "summary-table",
      htmltools::tags$caption(caption),
      htmltools::tags$thead(htmltools::tags$tr(lapply(headers, function(header) {
        htmltools::tags$th(scope = "col", header)
      }))),
      htmltools::tags$tbody(lapply(rows, function(row) {
        htmltools::tags$tr(
          htmltools::tags$th(scope = "row", row[[1]]),
          lapply(row[-1], htmltools::tags$td)
        )
      }))
    )
  ) %>% as.character() %>% cat("\n\n")
}

player_summary_table <- function(data, cell, caption) {
  rows <- lapply(difficulty_levels, function(difficulty) {
    c(list(difficulty), lapply(c("David", "Ryan"), function(player) {
      row <- filter(data, Difficulty == difficulty, Player == player)
      if (nrow(row) == 0) return("No recorded solves")
      cell(row)
    }))
  })
  summary_table(c("Game", "David", "Ryan"), rows, caption)
}

# Responsive plot convention for future sections:
# desktop = wide, side-by-side where useful; mobile portrait = 4.2 x ~7.3,
# wrapped titles, and legends below when a right legend would steal plot width.
# CSS switches phone landscape back to the desktop plot variant.
# Date plots use scale_month_axis() so every recorded month is labeled.
plot_solve_trend <- function(
  selected_difficulty,
  player_columns = 1,
  base_size = 14,
  title_width = Inf,
  subtitle_width = Inf
) {
  pips_data %>%
    filter(Difficulty == selected_difficulty) %>%
    ggplot(aes(x = Date, y = Solve_Time_Minutes, color = Difficulty)) +
    geom_point(alpha = 0.35, size = 1.6) +
    geom_smooth(method = "loess", span = 0.4, se = TRUE) +
    facet_wrap(~Player, ncol = player_columns) +
    scale_color_manual(values = difficulty_colors, drop = FALSE) +
    scale_month_axis() +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(
      title = chart_title(paste("Solve times:", selected_difficulty), title_width),
      subtitle = chart_title("David and Ryan share the same vertical scale", subtitle_width),
      x = "Date",
      y = "Solve time (minutes)"
    ) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold")
    )
}

plot_monthly_volatility <- function(
  selected_difficulty,
  player_columns = 1,
  base_size = 14,
  title_width = Inf,
  subtitle_width = Inf
) {
  pips_data %>%
    filter(Difficulty == selected_difficulty, Solve_Time_Minutes > 0) %>%
    ggplot(aes(x = Month, y = Solve_Time_Minutes, fill = Difficulty, group = Month)) +
    geom_boxplot(width = 22, alpha = 0.75, outlier.alpha = 0.55) +
    facet_wrap(~Player, ncol = player_columns) +
    scale_fill_manual(values = difficulty_colors, drop = FALSE) +
    scale_month_axis() +
    scale_y_log10(
      breaks = minute_breaks,
      labels = scales::label_number(accuracy = 0.1)
    ) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(
      title = chart_title(paste("Monthly consistency:", selected_difficulty), title_width),
      subtitle = chart_title("Shorter boxes mean less relative variation in solve times", subtitle_width),
      x = "Month",
      y = "Solve time (minutes, log scale)"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(lineheight = 0.9),
      plot.title = element_text(face = "bold")
    )
}

plot_player_distribution <- function(
  selected_difficulty,
  base_size = 14,
  title_width = Inf,
  subtitle_width = Inf
) {
  pips_data %>%
    filter(Difficulty == selected_difficulty, Solve_Time_Minutes > 0) %>%
    ggplot(aes(x = Player, y = Solve_Time_Minutes, fill = Player)) +
    geom_violin(alpha = 0.55, trim = FALSE) +
    geom_boxplot(width = 0.16, outlier.alpha = 0.35) +
    scale_fill_manual(values = c("David" = set1_colors[2], "Ryan" = set1_colors[1])) +
    scale_y_log10(
      breaks = minute_breaks,
      labels = scales::label_number(accuracy = 0.1)
    ) +
    theme_nyt_games(base_size = base_size) +
    labs(
      title = chart_title(paste("Solve-time distribution:", selected_difficulty), title_width),
      subtitle = chart_title("Wider sections show more solves at that time", subtitle_width),
      x = NULL,
      y = "Solve time (minutes, log scale)"
    ) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold")
    )
}

h2h_data <- overall_winners %>%
  filter(!is.na(Winner)) %>%
  arrange(Date) %>%
  mutate(
    David_Win = ifelse(Winner == "David", 1, 0),
    Rolling_Win_Rate = rollmean(David_Win, k = 10, fill = NA, align = "right")
  )

difficulty_h2h_data <- winner_data %>%
  filter(Difficulty %in% difficulty_levels, !is.na(Winner)) %>%
  mutate(Difficulty = factor(Difficulty, levels = difficulty_levels)) %>%
  arrange(Difficulty, Date) %>%
  group_by(Difficulty) %>%
  mutate(
    David_Win = ifelse(Winner == "David", 1, 0),
    Rolling_Win_Rate = rollmean(David_Win, k = 10, fill = NA, align = "right")
  ) %>%
  ungroup()

time_gap_data <- pips_data %>%
  select(Date, Player, Difficulty, Solve_Time_Minutes) %>%
  pivot_wider(names_from = Player, values_from = Solve_Time_Minutes) %>%
  filter(!is.na(David), !is.na(Ryan)) %>%
  mutate(
    Time_Gap_Minutes = David - Ryan,
    Faster_Player = case_when(
      Time_Gap_Minutes < 0 ~ "David faster",
      Time_Gap_Minutes > 0 ~ "Ryan faster",
      TRUE ~ "Same time"
    )
  )

overall_time_gap_data <- time_gap_data %>%
  group_by(Date) %>%
  summarise(Time_Gap_Minutes = sum(Time_Gap_Minutes, na.rm = TRUE), .groups = "drop") %>%
  mutate(Faster_Player = case_when(
    Time_Gap_Minutes < 0 ~ "David faster",
    Time_Gap_Minutes > 0 ~ "Ryan faster",
    TRUE ~ "Same time"
  ))

plot_h2h_win_rate <- function(
  selected_game = "Overall",
  base_size = 13,
  title_width = Inf,
  subtitle_width = Inf
) {
  is_overall <- selected_game == "Overall"
  plot_data <- if (is_overall) {
    h2h_data
  } else {
    difficulty_h2h_data %>% filter(Difficulty == selected_game)
  }
  line_color <- if (is_overall) set1_colors[2] else difficulty_colors[[selected_game]]

  ggplot(plot_data, aes(x = Date, y = Rolling_Win_Rate)) +
    geom_line(color = line_color, linewidth = 1.2, na.rm = TRUE) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = set1_colors[1]) +
    scale_month_axis() +
    scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(
      title = chart_title(paste("10-result win rate:", selected_game), title_width),
      subtitle = chart_title("David's wins in the last 10 recorded results; ties count as non-wins", subtitle_width),
      x = "Date",
      y = "David's win rate (%)"
    ) +
    theme(plot.title = element_text(face = "bold"))
}

plot_time_gap <- function(
  selected_game = "Overall",
  base_size = 13,
  title_width = Inf,
  subtitle_width = Inf,
  legend_position = "right"
) {
  is_overall <- selected_game == "Overall"
  plot_data <- if (is_overall) {
    overall_time_gap_data
  } else {
    time_gap_data %>% filter(Difficulty == selected_game)
  }

  ggplot(plot_data, aes(x = Date, y = Time_Gap_Minutes, color = Faster_Player)) +
    geom_hline(yintercept = 0, color = "grey35") +
    geom_point(alpha = 0.5, size = 1.5) +
    geom_smooth(aes(group = 1), method = "loess", span = 0.45, se = FALSE, color = "black") +
    scale_color_manual(values = c(
      "David faster" = set1_colors[2],
      "Ryan faster" = set1_colors[1],
      "Same time" = "#7f7f7f"
    )) +
    guides(color = guide_legend(
      nrow = if (legend_position == "bottom") 2 else 3,
      byrow = TRUE
    )) +
    scale_month_axis() +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(
      title = chart_title(paste("Daily time gap:", selected_game), title_width),
      subtitle = chart_title("Negative values favor David; positive values favor Ryan", subtitle_width),
      x = "Date",
      y = "David minus Ryan (minutes)",
      color = if (legend_position == "bottom") NULL else "Faster player"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = legend_position,
      legend.direction = if_else(legend_position == "bottom", "horizontal", "vertical"),
      legend.box = "horizontal"
    )
}

plot_monthly_trend <- function(
  selected_difficulty,
  player_columns = 2,
  base_size = 13,
  title_width = Inf,
  subtitle_width = Inf
) {
  monthly_trend_data %>%
    filter(Difficulty == selected_difficulty, !is.na(Change_Minutes)) %>%
    ggplot(aes(x = Month, y = Change_Minutes, fill = Player, alpha = Is_Partial)) +
    geom_hline(yintercept = 0, color = nyt_black, linewidth = 0.5) +
    geom_col(width = 22) +
    geom_point(
      data = function(data) filter(data, near(Change_Minutes, 0)),
      shape = 21,
      size = 2.5,
      stroke = 0.5,
      color = nyt_black
    ) +
    facet_wrap(~Player, ncol = player_columns) +
    scale_fill_manual(values = c("David" = set1_colors[2], "Ryan" = set1_colors[1])) +
    scale_alpha_manual(values = c("FALSE" = 0.85, "TRUE" = 0.45)) +
    scale_month_axis() +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(
      title = chart_title(paste("Monthly pace change:", selected_difficulty), title_width),
      subtitle = chart_title("Below zero is faster than the previous month; above zero is slower", subtitle_width),
      caption = chart_title("Dots mean no change. Lighter bars mark a partial month.", subtitle_width),
      x = "Month",
      y = "Change in median (minutes)"
    ) +
    theme(legend.position = "none")
}

source("relative_performance.R")
hard_source <- pips_data %>%
  filter(Difficulty == "Hard") %>%
  select(Date, Player, Recorded_Solve_Time_Seconds) %>%
  pivot_wider(names_from = Player, values_from = Recorded_Solve_Time_Seconds,
    names_glue = "{Player}_Seconds")
relative_data <- prepare_relative_pairs(as.data.frame(hard_source),
  cap_seconds = hard_time_cap_minutes * 60, as_of = today(tzone = "America/New_York"))
hard_pairs <- relative_data$pairs
relative_analysis <- analyze_relative_performance(hard_pairs)
relative_primary <- relative_analysis$primary
hard_recorded_winners <- winner_data %>% filter(Difficulty == "Hard") %>% select(Date, Winner)
hard_outcome_check <- as_tibble(hard_pairs) %>% left_join(hard_recorded_winners, by = "Date")
write.csv(hard_pairs, "relative-performance-data.csv", row.names = FALSE)

signed_percent <- function(value, accuracy = 0.01) {
  paste0(ifelse(value > 0, "+", ""), scales::percent(value, accuracy = accuracy))
}

relative_conclusion <- if (relative_primary$lower <= 0 && relative_primary$upper >= 0) {
  if (relative_primary$slope > 0) {
    "The estimated trend points toward David taking longer relative to Ryan, but the evidence does not establish a sustained decline."
  } else {
    "The estimated trend points toward David taking less time relative to Ryan, but the evidence does not establish a sustained improvement."
  }
} else if (relative_primary$lower > 0) {
  "The fitted trend supports David taking longer relative to Ryan over this record, under the model's assumptions."
} else {
  "The fitted trend supports David taking less time relative to Ryan over this record, under the model's assumptions."
}

plot_relative_performance <- function(base_size = 13, title_width = Inf,
  subtitle_width = Inf) {
  ggplot() +
    geom_hline(yintercept = 1, linetype = "dashed", color = nyt_gray) +
    geom_ribbon(data = relative_analysis$prediction,
      aes(x = Date, ymin = Lower, ymax = Upper), fill = set1_colors[2], alpha = 0.15) +
    geom_point(data = hard_pairs, aes(x = Date, y = exp(Log_Ratio), color = "Daily ratio"),
      size = 1.3, alpha = 0.35) +
    geom_line(data = relative_analysis$prediction, aes(x = Date, y = Ratio),
      color = nyt_black, linewidth = 1) +
    geom_point(data = relative_analysis$monthly,
      aes(x = Month, y = Ratio, color = "Monthly geometric mean"), size = 2.8) +
    scale_color_manual(values = c("Daily ratio" = nyt_gray,
      "Monthly geometric mean" = set1_colors[1])) +
    scale_month_axis() +
    scale_y_log10(breaks = c(0.125, 0.25, 0.5, 1, 2, 4, 8, 16),
      labels = function(values) paste0(values, "x")) +
    theme_nyt_games(base_size = base_size) +
    theme_month_axis(is_mobile = is.finite(title_width)) +
    labs(title = chart_title("Hard time ratios", title_width),
      subtitle = chart_title("Above 1x means David took longer; below 1x means Ryan took longer", subtitle_width),
      x = "Date", y = "David / Ryan time ratio (log scale)", color = NULL) +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    theme(legend.position = "bottom", legend.direction = "horizontal",
      legend.box = "horizontal")
}
```

The report includes recorded solves from `r format(min(pips_data$Date), '%B %d, %Y')` through `r format(max(pips_data$Date), '%B %d, %Y')`. Missing times are excluded. Solve times are shown in minutes. The median represents a typical solve time and is less affected by very slow solves than the average. Table values are rounded for display; changes are calculated from unrounded medians.

Hard times above `r hard_time_cap_minutes` minutes are treated as `r hard_time_cap_minutes` minutes in every solve-time analysis and in the data table. This cap reduces the measured spread and can make different Hard times appear equal. A capped time doesn't necessarily mean the puzzle was completed. Win rates use the winners recorded in the sheet, not the capped times.

The current month's summaries use only the results recorded so far. Comparisons with a full month can change as more results are added.
    
### 1. Solve times over time

Each dot is a recorded solve. The line smooths nearby results to show the general direction: lower means faster. The shaded band shows uncertainty in the fitted line. David and Ryan share the same vertical scale within each game. Puzzle difficulty and other conditions can affect the trend.
    
::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Easy

```{r trend-easy-desktop}
#| fig-width: 10
#| fig-height: 5
plot_solve_trend("Easy", player_columns = 2, base_size = 13)
```

#### Medium

```{r trend-medium-desktop}
#| fig-width: 10
#| fig-height: 5
plot_solve_trend("Medium", player_columns = 2, base_size = 13)
```

#### Hard

```{r trend-hard-desktop}
#| fig-width: 10
#| fig-height: 5
plot_solve_trend("Hard", player_columns = 2, base_size = 13)
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Easy

```{r trend-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_solve_trend("Easy", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Medium

```{r trend-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_solve_trend("Medium", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Hard

```{r trend-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_solve_trend("Hard", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

:::

:::

```{r trend-interpretation, results='asis'}
trend_summary <- pips_data %>%
  arrange(Player, Difficulty, Date) %>%
  group_by(Player, Difficulty) %>%
  summarise(
    first_30_median = median(head(Solve_Time_Minutes, 30), na.rm = TRUE),
    recent_30_median = median(tail(Solve_Time_Minutes, 30), na.rm = TRUE),
    change_minutes = recent_30_median - first_30_median,
    comparison_solves = min(n(), 30),
    .groups = "drop"
  )

cat("The table compares each player's first and latest 30 recorded solves. These medians are separate from the fitted line.\n\n")
player_summary_table(trend_summary, function(row) {
  summary_cell(
    pace_change(row$change_minutes),
    paste("First", row$comparison_solves, "median:", format_minutes(row$first_30_median)),
    paste("Latest", row$comparison_solves, "median:", format_minutes(row$recent_30_median))
  )
}, "Table 1. Change in median solve time")
```

### 2. Monthly consistency

Each box spans the middle half of a month's solve times. The line inside marks the median. A shorter box means those times were closer together; it doesn't necessarily mean faster solves. Points beyond the whiskers show unusually high or low times. The logarithmic vertical scale gives equal space to equal ratios, such as 1 to 2 minutes and 2 to 4 minutes.

::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Easy

```{r volatility-easy-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_volatility("Easy", player_columns = 2, base_size = 13)
```

#### Medium

```{r volatility-medium-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_volatility("Medium", player_columns = 2, base_size = 13)
```

#### Hard

```{r volatility-hard-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_volatility("Hard", player_columns = 2, base_size = 13)
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Easy

```{r volatility-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_volatility("Easy", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Medium

```{r volatility-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_volatility("Medium", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Hard

```{r volatility-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_volatility("Hard", title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

:::

:::

```{r volatility-interpretation, results='asis'}
volatility_summary <- pips_data %>%
  group_by(Player, Difficulty, Month) %>%
  summarise(
    median_minutes = median(Solve_Time_Minutes, na.rm = TRUE),
    iqr_minutes = IQR(Solve_Time_Minutes, na.rm = TRUE),
    solves = n(),
    .groups = "drop"
  ) %>%
  arrange(Player, Difficulty, Month) %>%
  group_by(Player, Difficulty) %>%
  summarise(
    typical_iqr = if (any(Month < report_month)) median(iqr_minutes[Month < report_month]) else NA_real_,
    latest_month = last(Month),
    latest_iqr = last(iqr_minutes),
    latest_median = last(median_minutes),
    latest_solves = last(solves),
    .groups = "drop"
  )

cat("The table compares the latest month's spread with the median spread of earlier calendar months that have ended. Spread is the interquartile range: the range containing the middle half of recorded times, measured in minutes before applying the logarithmic scale.\n\n")
player_summary_table(volatility_summary, function(row) {
  comparison <- case_when(
    is.na(row$typical_iqr) ~ "No earlier-month comparison",
    near(row$latest_iqr, row$typical_iqr) ~ "Same spread as earlier months",
    row$latest_iqr < row$typical_iqr ~ "Less spread than earlier months",
    TRUE ~ "More spread than earlier months"
  )
  summary_cell(
    comparison,
    paste(format(row$latest_month, "%B %Y"), "spread:", format_minutes(row$latest_iqr)),
    paste("Earlier-month benchmark:", format_minutes(row$typical_iqr)),
    paste(row$latest_solves, "solves;", if (row$latest_month >= report_month) "partial month" else "calendar month ended")
  )
}, "Table 2. Monthly variation in solve times")
```
    
### 3. Solve-time distribution

The shapes summarize all recorded times, without showing when they occurred. Wider sections mean more solves near that time. The box inside each shape shows the middle half of times and the median. The vertical axis is logarithmic, as in the monthly consistency chart. Each player's summary includes their own recorded solves, so the medians aren't restricted to matched days.

::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Easy

```{r player-distribution-easy-desktop}
#| fig-width: 10
#| fig-height: 4.6
plot_player_distribution("Easy", base_size = 13)
```

#### Medium

```{r player-distribution-medium-desktop}
#| fig-width: 10
#| fig-height: 4.6
plot_player_distribution("Medium", base_size = 13)
```

#### Hard

```{r player-distribution-hard-desktop}
#| fig-width: 10
#| fig-height: 4.6
plot_player_distribution("Hard", base_size = 13)
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Easy

```{r player-distribution-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_player_distribution("Easy", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Medium

```{r player-distribution-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_player_distribution("Medium", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Hard

```{r player-distribution-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_player_distribution("Hard", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

:::

:::

```{r player-distribution-interpretation, results='asis'}
distribution_summary <- pips_data %>%
  group_by(Difficulty, Player) %>%
  summarise(
    median_minutes = median(Solve_Time_Minutes, na.rm = TRUE),
    lower_quartile = quantile(Solve_Time_Minutes, 0.25, na.rm = TRUE),
    upper_quartile = quantile(Solve_Time_Minutes, 0.75, na.rm = TRUE),
    solves = n(),
    .groups = "drop"
  )

cat("The table compares typical solve times and the range containing the middle half of each player's results.\n\n")
player_summary_table(distribution_summary, function(row) {
  summary_cell(
    paste("Median:", format_minutes(row$median_minutes)),
    paste("Middle half:", format_minutes(row$lower_quartile), "to", format_minutes(row$upper_quartile)),
    paste(row$solves, "recorded solves")
  )
}, "Table 3. Solve-time medians and ranges")
```

### 4. Monthly median solve times

Each cell shows a month's median solve time in minutes. The color identifies the game, and darker shading means a higher median. Use the printed values for precise comparisons, especially for Easy, where small changes can be hard to see in the shading. The current month's medians are provisional.

```{r monthly-heatmap}
#| fig-width: 10
#| fig-height: 6
monthly_medians <- pips_data %>%
  group_by(Player, Month, Difficulty) %>%
  summarise(Median_Minutes = median(Solve_Time_Minutes, na.rm = TRUE), .groups = "drop")

ggplot(monthly_medians, aes(x = Month, y = Difficulty, fill = Difficulty, alpha = Median_Minutes)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = scales::number(Median_Minutes, accuracy = 0.01)), size = 3) +
  facet_wrap(~Player, ncol = 1) +
  scale_fill_brewer(palette = "Set1", drop = FALSE) +
  scale_alpha_continuous(range = c(0.35, 1), name = "Median minutes") +
  scale_month_axis() +
  theme_nyt_games() +
  theme_month_axis() +
  labs(
    title = "Monthly median solve times",
    subtitle = "Darker cells indicate slower monthly medians",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

```{r monthly-heatmap-interpretation, results='asis'}
monthly_summary <- monthly_medians %>%
  arrange(Player, Difficulty, Month) %>%
  group_by(Player, Difficulty) %>%
  summarise(
    best_months = paste(format(Month[near(Median_Minutes, min(Median_Minutes))], "%b %Y"), collapse = ", "),
    best_median = min(Median_Minutes, na.rm = TRUE),
    latest_month = last(Month),
    latest_median = last(Median_Minutes),
    .groups = "drop"
  )

cat("The table lists the lowest monthly median and the latest month's median. A lowest median based on a partial month isn't a full-month record.\n\n")
player_summary_table(monthly_summary, function(row) {
  summary_cell(
    paste("Lowest median:", format_minutes(row$best_median)),
    paste("Month(s):", row$best_months),
    paste(format(row$latest_month, "%B %Y"), "median:", format_minutes(row$latest_median)),
    if (row$latest_month >= report_month) "Latest month is partial" else "Latest calendar month has ended"
  )
}, "Table 4. Lowest and latest monthly medians")
```

### 5. Month-to-month pace change

Each bar shows the change in median solve time from the previous calendar month. Below zero means faster; above zero means slower. A dot at zero means no change, not missing data. A month without data in the immediately preceding month has no comparison. Lighter bars mark the current, partial month. David and Ryan share the same vertical scale within each game.

::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Easy

```{r monthly-trend-easy-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_trend("Easy")
```

#### Medium

```{r monthly-trend-medium-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_trend("Medium")
```

#### Hard

```{r monthly-trend-hard-desktop}
#| fig-width: 10
#| fig-height: 5
plot_monthly_trend("Hard")
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Easy

```{r monthly-trend-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_trend("Easy", player_columns = 1, base_size = 14, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Medium

```{r monthly-trend-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_trend("Medium", player_columns = 1, base_size = 14, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Hard

```{r monthly-trend-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
plot_monthly_trend("Hard", player_columns = 1, base_size = 14, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

:::

:::

```{r monthly-trend-interpretation, results='asis'}
latest_monthly_trends <- monthly_trend_data %>%
  group_by(Player, Difficulty) %>%
  slice_max(Month, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(Difficulty, Player)

cat("Each player's latest month is compared with the previous calendar month in the table. Partial-month results can change substantially as more solves are recorded.\n\n")
player_summary_table(latest_monthly_trends, function(row) {
  percent_text <- if (!is.na(row$Change_Percent) && !near(row$Change_Minutes, 0)) {
    paste0(" (", scales::percent(abs(row$Change_Percent), accuracy = 0.1), ")")
  } else ""
  previous_text <- if (is.na(row$Previous_Median)) {
    "Previous calendar month: no recorded solves"
  } else paste(format(row$Previous_Month, "%B %Y"), "median:", format_minutes(row$Previous_Median))
  summary_cell(
    paste0(pace_change(row$Change_Minutes), percent_text),
    paste(format(row$Month, "%B %Y"), "median:", format_minutes(row$Median_Minutes)),
    previous_text,
    paste(row$Solves, "solves through", format(row$Last_Date, "%B %d"), if (row$Is_Partial) "(partial month)" else "")
  )
}, "Table 5. Latest month compared with the previous month")
```

### 6. Head-to-head win rate

A win rate summarizes head-to-head results without measuring the size of the time difference. The line shows David's share of wins in the last 10 recorded results. Above 50% means David won more than half of those results. A tie counts as a non-win for both players, so a value below 50% doesn't always mean Ryan won a majority. The Overall tab uses the sheet's daily overall winner; the other tabs use the winner for that game. This is a recent win rate, not a forecast.
    
::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Overall

```{r win-rate-overall-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_h2h_win_rate("Overall", base_size = 13)
```

#### Easy

```{r win-rate-easy-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_h2h_win_rate("Easy", base_size = 13)
```

#### Medium

```{r win-rate-medium-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_h2h_win_rate("Medium", base_size = 13)
```

#### Hard

```{r win-rate-hard-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_h2h_win_rate("Hard", base_size = 13)
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Overall

```{r win-rate-overall-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_h2h_win_rate("Overall", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Easy

```{r win-rate-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_h2h_win_rate("Easy", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Medium

```{r win-rate-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_h2h_win_rate("Medium", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

#### Hard

```{r win-rate-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_h2h_win_rate("Hard", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width)
```

:::

:::

```{r win-rate-interpretation, results='asis'}
win_record <- bind_rows(
  mutate(h2h_data, Game = "Overall"),
  mutate(difficulty_h2h_data, Game = as.character(Difficulty))
) %>%
  mutate(Game = factor(Game, levels = c("Overall", difficulty_levels))) %>%
  arrange(Game, Date) %>%
  group_by(Game) %>%
  summarise(
    david_wins = sum(Winner == "David", na.rm = TRUE),
    ryan_wins = sum(Winner == "Ryan", na.rm = TRUE),
    ties = sum(Winner == "Tie", na.rm = TRUE),
    results = n(),
    david_rate = mean(David_Win, na.rm = TRUE),
    latest_rolling = last(Rolling_Win_Rate),
    latest_date = last(Date),
    recent_david_wins = sum(tail(Winner, 10) == "David"),
    recent_ryan_wins = sum(tail(Winner, 10) == "Ryan"),
    recent_ties = sum(tail(Winner, 10) == "Tie"),
    .groups = "drop"
  )

cat("The full record and the last 10 results are shown separately in the table. Ties stay in the denominator.\n\n")
win_rows <- lapply(seq_len(nrow(win_record)), function(i) {
  row <- win_record[i, ]
  list(
    as.character(row$Game),
    summary_cell(
      paste("David:", row$david_wins, "wins; Ryan:", row$ryan_wins, "wins"),
      paste("Ties:", row$ties, "of", row$results, "results"),
      paste("David's win rate:", scales::percent(row$david_rate, accuracy = 0.1))
    ),
    if (is.na(row$latest_rolling)) "Fewer than 10 recorded results" else summary_cell(
      paste("David's win rate:", scales::percent(row$latest_rolling, accuracy = 1)),
      paste("David:", row$recent_david_wins, "wins; Ryan:", row$recent_ryan_wins, "wins; ties:", row$recent_ties),
      paste("Through", format(row$latest_date, "%B %d, %Y"))
    )
  )
})
summary_table(c("Game", "All recorded results", "Last 10 results"), win_rows, "Table 6. Recorded wins and recent win rates")
```

### 7. Daily time gap

A small time advantage and a large one count equally as wins. The daily gap shows the size of the difference: David's time minus Ryan's time for the same date and game. Negative values mean David was faster; positive values mean Ryan was faster. Zero means equal analyzed times, including ties created by the Hard cap. Only games with both players' times are included. The Overall tab adds the gaps for the matched games on each date; a day with missing times can include fewer than three games.

::: {.responsive-plot-desktop}

::: {.panel-tabset}

#### Overall

```{r time-gap-overall-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_time_gap("Overall", base_size = 13)
```

#### Easy

```{r time-gap-easy-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_time_gap("Easy", base_size = 13)
```

#### Medium

```{r time-gap-medium-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_time_gap("Medium", base_size = 13)
```

#### Hard

```{r time-gap-hard-desktop}
#| fig-width: 10
#| fig-height: 4.8
plot_time_gap("Hard", base_size = 13)
```

:::

:::

::: {.responsive-plot-mobile}

::: {.panel-tabset}

#### Overall

```{r time-gap-overall-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_time_gap("Overall", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width, legend_position = "bottom")
```

#### Easy

```{r time-gap-easy-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_time_gap("Easy", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width, legend_position = "bottom")
```

#### Medium

```{r time-gap-medium-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_time_gap("Medium", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width, legend_position = "bottom")
```

#### Hard

```{r time-gap-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.2
plot_time_gap("Hard", base_size = 15, title_width = mobile_title_width, subtitle_width = mobile_subtitle_width, legend_position = "bottom")
```

:::

:::

```{r time-gap-interpretation, results='asis'}
gap_summary <- bind_rows(
  transmute(overall_time_gap_data, Game = "Overall", Time_Gap_Minutes),
  transmute(time_gap_data, Game = as.character(Difficulty), Time_Gap_Minutes)
) %>%
  mutate(Game = factor(Game, levels = c("Overall", difficulty_levels))) %>%
  group_by(Game) %>%
  summarise(
    median_gap = median(Time_Gap_Minutes, na.rm = TRUE),
    david_faster_share = mean(Time_Gap_Minutes < 0, na.rm = TRUE),
    ryan_faster_share = mean(Time_Gap_Minutes > 0, na.rm = TRUE),
    same_time_share = mean(Time_Gap_Minutes == 0, na.rm = TRUE),
    matched_days = n(),
    .groups = "drop"
  )

cat("The table summarizes matched days across the full record. Equal times are counted separately from either player's faster days.\n\n")
gap_rows <- lapply(seq_len(nrow(gap_summary)), function(i) {
  row <- gap_summary[i, ]
  list(
    as.character(row$Game),
    if (near(row$median_gap, 0)) "No median daily gap" else paste(
      if (row$median_gap < 0) "David faster by" else "Ryan faster by",
      format_minutes(abs(row$median_gap), 0.001)
    ),
    summary_cell(
      paste("David faster:", scales::percent(row$david_faster_share, accuracy = 0.1)),
      paste("Ryan faster:", scales::percent(row$ryan_faster_share, accuracy = 0.1)),
      paste("Equal times:", scales::percent(row$same_time_share, accuracy = 0.1)),
      paste(row$matched_days, "matched days")
    )
  )
})
summary_table(c("Game", "Median daily gap", "Share of matched days"), gap_rows, "Table 7. Same-day solve-time comparisons")
```

### 8. Is relative Hard performance changing?

The daily gap shows who took longer on a shared puzzle. This analysis asks a different question: has David's time **relative to Ryan's** changed over the full record? It uses only Hard puzzles with a positive recorded time for both players, capped at 20 minutes each.

```{r relative-primary-summary, results='asis'}
cat("**", relative_conclusion, "**\n\n", sep = "")
cat("The test includes ", nrow(hard_pairs), " shared Hard puzzles from ",
  format(min(hard_pairs$Date), "%B %d, %Y"), " through ",
  format(max(hard_pairs$Date), "%B %d, %Y"),
  ". Days without both times are excluded.\n\n", sep = "")
summary_table(c("Primary result", "Estimate"), list(
  list("Change in David/Ryan time ratio per 30 days", signed_percent(expm1(relative_primary$slope))),
  list("95% confidence interval", paste(signed_percent(expm1(relative_primary$lower)), "to", signed_percent(expm1(relative_primary$upper)))),
  list("Two-sided p-value", scales::number(relative_primary$p_value, accuracy = 0.001))
), "Table 8. Full-record relative-time trend", statistical = TRUE)
```

For day $d$, let $D_d$ and $R_d$ be David's and Ryan's recorded times in seconds. The outcome is the log of the capped time ratio:

$$
y_d = \ln\!\left(\frac{\min(D_d,1200)}{\min(R_d,1200)}\right).
$$

The test fits a straight-line trend in this log ratio over calendar time. A positive slope means David's relative time increased; a negative slope means it decreased. The reported percentage is $100(e^{\beta}-1)$, where $\beta$ is the change in log ratio per 30 days. It is a proportional change in the time ratio, not a change in win probability.

::: {.responsive-plot-desktop}

```{r relative-hard-desktop}
#| fig-width: 10
#| fig-height: 5.4
#| fig-alt: "David-to-Ryan capped Hard time ratios by date, with monthly geometric means, a fitted trend, and its pointwise 95% confidence band. A ratio above one means David took longer."
plot_relative_performance()
```

:::

::: {.responsive-plot-mobile}

```{r relative-hard-mobile}
#| fig-width: 4.2
#| fig-height: 7.4
#| fig-alt: "David-to-Ryan capped Hard time ratios by date, with monthly geometric means, a fitted trend, and its pointwise 95% confidence band. A ratio above one means David took longer."
plot_relative_performance(base_size = 14, title_width = mobile_title_width,
  subtitle_width = mobile_subtitle_width)
```

:::

A ratio of 1 means equal capped times. Above 1 means David took longer; below 1 means Ryan took longer. The black line is the fitted trend. Its shaded band is a pointwise 95% confidence interval for the fitted mean log ratio, converted back to ratios. It is not a prediction interval for the next puzzle. The monthly points are geometric means of the individual puzzle ratios, not ratios of the players' separate monthly medians.

```{r relative-interpretation, results='asis'}
if (relative_primary$lower <= 0 && relative_primary$upper >= 0) {
  cat("The confidence interval includes improvement, no change, and deterioration. The result is inconclusive; it does not establish that performance stayed unchanged.\n\n")
} else {
  cat("The confidence interval excludes no change under this model. That supports a trend across the recorded period, not a guarantee that it will continue.\n\n")
}
cat("The p-value is not the probability that David is getting worse or that there is no trend. It describes how unusual an estimated slope this far from zero would be if the no-trend hypothesis and model assumptions held.\n\n")
```

#### How sensitive is the conclusion?

The following checks change the dependence window or the dates included. The full-record, seven-day test remains the main comparison. The April and recent-90-day windows are exploratory; they are not independent confirmations of the main result.

```{r relative-sensitivity, results='asis'}
checks <- relative_analysis$checks
check_rows <- lapply(seq_len(nrow(checks)), function(i) {
  row <- checks[i, ]
  label <- row$Check
  if (label == "Exclude latest calendar month") {
    label <- paste("Exclude", format(max(hard_pairs$Date), "%B %Y"))
  }
  list(label, summary_cell(
    paste(signed_percent(row$Change), "per 30 days"),
    paste("95% CI:", signed_percent(row$Lower), "to", signed_percent(row$Upper)),
    paste("p =", scales::number(row$P, accuracy = 0.001), ";", row$Pairs, "shared puzzles")
  ))
})
bootstrap <- relative_analysis$bootstrap
check_rows <- c(check_rows, list(list("Seven-game block bootstrap", summary_cell(
  paste("95% CI:", signed_percent(expm1(bootstrap$lower)), "to", signed_percent(expm1(bootstrap$upper))),
  paste(scales::comma(bootstrap$replicates), "resamples; fixed calendar dates"),
  "Supporting uncertainty check; no separate p-value"
))))
summary_table(c("Check", "Relative-time result"), check_rows,
  "Table 9. Sensitivity of the estimated relative-time trend", statistical = TRUE)
all_include_zero <- all(checks$Lower <= 0 & checks$Upper >= 0) &&
  bootstrap$lower <= 0 && bootstrap$upper >= 0
cat(if (all_include_zero) {
  "Every interval above includes no change. These checks do not establish a sustained relative-time trend.\n\n"
} else {
  "The intervals do not all reach the same conclusion. Treat the differences as sensitivity to the method or date window, not as a reason to select the lowest p-value.\n\n"
})
```

The supporting win test recomputes winners from capped times and excludes ties. It therefore differs from the earlier win-rate charts, which use the sheet's recorded winners. A logistic regression estimates the change in David's odds of winning per 30 days. An odds multiplier below 1 means declining odds, not a fixed percentage-point drop in win rate.

```{r relative-supporting-tests, results='asis'}
wins <- relative_analysis$wins
gap <- relative_analysis$time_gap
summary_table(c("Supporting test", "Estimate"), list(
  list("David's win odds per 30 days", summary_cell(
    paste0(scales::number(exp(wins$slope), accuracy = 0.001), "x"),
    paste("95% CI:", scales::number(exp(wins$lower), accuracy = 0.001), "to", scales::number(exp(wins$upper), accuracy = 0.001)),
    paste("p =", scales::number(wins$p_value, accuracy = 0.001), ";", wins$n, "decisive games")
  )),
  list("David-minus-Ryan gap per 30 days", summary_cell(
    paste0(if (gap$slope > 0) "+" else "", format_minutes(gap$slope, 0.001)),
    paste("95% CI:", format_minutes(gap$lower, 0.001), "to", format_minutes(gap$upper, 0.001)),
    paste("p =", scales::number(gap$p_value, accuracy = 0.001), ";", gap$n, "shared puzzles")
  ))
), "Table 10. Trends in capped-time wins and daily gaps", statistical = TRUE)
if (wins$lower <= 0 && wins$upper >= 0) {
  cat("The win-odds interval includes 1, so this supporting test is also inconclusive about a trend.\n\n")
}
```

#### What does the history show?

These periods summarize the observed puzzle ratios; they are not separate trend tests. A geometric mean near 1 indicates similar times, while a value above 1 indicates that David took longer. The latest period can contain only a few shared puzzles.

```{r relative-periods, results='asis'}
periods <- relative_analysis$periods
period_rows <- lapply(seq_len(nrow(periods)), function(i) {
  row <- periods[i, ]
  list(row$Period, row$Pairs, paste0(scales::number(row$Ratio, accuracy = 0.001), "x"))
})
summary_table(c("Period", "Shared puzzles", "David/Ryan time ratio"), period_rows,
  "Table 11. Geometric mean of within-puzzle time ratios", statistical = TRUE)
latest <- tail(relative_analysis$monthly, 1)
cat(format(latest$Month, "%B %Y"), " has ", latest$Pairs,
  " shared puzzles through ", format(max(hard_pairs$Date), "%B %d"),
  ", with a geometric mean ratio of ", scales::number(latest$Ratio, accuracy = 0.001),
  "x. This describes those games; it does not establish a lasting change.\n\n", sep = "")
if (nrow(periods) >= 3 && abs(log(periods$Ratio[3])) < abs(log(periods$Ratio[2]))) {
  cat("July-September was closer to time parity than April-June. That recovery makes a simple story of steadily losing ground a poor description of this record.\n\n")
}
```

<details>
<summary>Monthly paired results</summary>

The table uses only shared Hard puzzles. Winners are recomputed after capping; ties are listed separately.

```{r relative-monthly-results, results='asis'}
monthly <- relative_analysis$monthly
monthly_rows <- lapply(seq_len(nrow(monthly)), function(i) {
  row <- monthly[i, ]
  list(format(row$Month, "%b %Y"), summary_cell(
    paste(row$Pairs, "shared puzzles"),
    paste("David:", row$David_Wins, "wins; Ryan:", row$Ryan_Wins, "wins; ties:", row$Ties)
  ), summary_cell(
    paste0(scales::number(row$Ratio, accuracy = 0.001), "x geometric mean"),
    paste0(scales::number(row$Median_Ratio, accuracy = 0.001), "x median ratio")
  ))
})
summary_table(c("Month", "Capped-time results", "David/Ryan ratio"), monthly_rows,
  "Table 12. Monthly relative Hard results", statistical = TRUE)
```

</details>

<details>
<summary>Methods, assumptions, and data checks</summary>

The model fits $y_d = \alpha + \beta t_d + \epsilon_d$, where $t_d$ is calendar days since the first shared puzzle, divided by 30. The null hypothesis is $\beta = 0$; the test is two-sided. Confidence intervals use a normal reference distribution, so the inference is approximate and depends on the model's assumptions.

Newey-West uncertainty estimates allow unequal residual variation and short-term dependence. The main test uses Bartlett weights, a seven-calendar-day lag window, and an $n/(n-2)$ finite-sample correction. Missing dates contribute zero score vectors to the covariance calculation, not invented solve times or fitted observations. The win-odds model uses the same calendar-aware covariance method.

The bootstrap uses 10,000 circular residual resamples, blocks of seven successive shared games, seed 20261005, and a basic 95% interval. Calendar dates stay fixed. It assumes the residual process is approximately stationary around the fitted trend. R's random-number generator differs from the attachment's Python implementation, so bootstrap endpoints can differ slightly even with the same seed.

Comparing two times for the same puzzle reduces the effect of shared difficulty **if it changes both times proportionally**. Puzzle styles can still favor one player, and capping compresses large differences. Relative results cannot distinguish Ryan improving from David declining. These are retrospective analyses, not preregistered tests; repeated checks and selected date windows can give misleadingly small p-values.

```{r relative-data-checks, results='asis'}
excluded_dates <- relative_data$excluded$Date
audit_rows <- list(
  list("Paired records used", nrow(hard_pairs)),
  list("Hard dates excluded", if (length(excluded_dates)) paste(format(sort(excluded_dates), "%b %d, %Y"), collapse = "; ") else "None"),
  list("Recorded times above 20 minutes", paste("David:", sum(hard_pairs$David_Was_Capped), "; Ryan:", sum(hard_pairs$Ryan_Was_Capped))),
  list("Analyzed times at the cap", paste("David:", sum(hard_pairs$David_Seconds == 1200), "; Ryan:", sum(hard_pairs$Ryan_Seconds == 1200))),
  list("Both players at the cap", sum(hard_pairs$David_Seconds == 1200 & hard_pairs$Ryan_Seconds == 1200)),
  list("Capped-time outcomes", paste("David:", sum(hard_pairs$Capped_Winner == "David"), "wins; Ryan:", sum(hard_pairs$Capped_Winner == "Ryan"), "wins; ties:", sum(hard_pairs$Capped_Winner == "Tie"))),
  list("Outcomes changed by capping", sum(hard_outcome_check$Winner != hard_outcome_check$Capped_Winner, na.rm = TRUE))
)
summary_table(c("Data check", "Result"), audit_rows,
  "Table 13. Paired Hard data checks", statistical = TRUE)
```

The source sheet is unchanged. The downloaded data contain capped times, not the original over-cap durations. [Paired data (CSV)](relative-performance-data.csv) and [analysis code (R)](relative_performance.R) reproduce the tests without Google credentials:

```r
source("relative_performance.R")
pairs <- read.csv("relative-performance-data.csv")
results <- analyze_relative_performance(pairs)
results$checks
```

Method references: [Newey-West covariance](https://www.statsmodels.org/stable/generated/statsmodels.stats.sandwich_covariance.cov_hac.html), [reference covariance implementation](https://www.statsmodels.org/stable/_modules/statsmodels/stats/sandwich_covariance.html), and [R's block-bootstrap documentation](https://stat.ethz.ch/R-manual/R-devel/library/boot/html/tsboot.html). Regression uses R's `lm()` and `glm()`; block resampling uses `boot::tsboot()`.

For the next review, keep the same ratio and cap, choose a future review date before examining the new results, and assess the confidence interval then. This avoids treating each disappointing game as a new statistical verdict.

</details>

### 9. Recorded results

This table lists the analyzed solve times and the winners recorded in the sheet. Hard times are capped at 20 minutes here, just as they are in the charts. A recorded winner can differ from the faster analyzed time when the cap creates a tie.

```{r data-table}
pips_data %>%
  arrange(desc(Date)) %>%
  select(Date, Player, Difficulty, Solve_Time_Minutes, Winner) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    colnames = c("Date", "Player", "Game", "Solve time (min)", "Recorded winner"),
    rownames = FALSE,
    class = 'cell-border stripe'
  ) %>%
  formatRound("Solve_Time_Minutes", digits = 2)
```
